---
title: 'Section 4: Data Wrangling (Part 2)'
author: 'MTH 365: Introduction to Data Science'
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document
---

## Recommended Reading

- _Modern Data Science with R_ Ch. 4: Data Wrangling
- _Modern Data Science with R_ Ch. 5: Tidy Data and Iteration
- Wickham, Hadley. (2014). "Tidy Data". _Journal of Statistical Software_ 59(10). Available on BlueLine.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

## Example: NYC flights

The R package `nycflights13` contains data about all flights that departed one of the three New York City airports (JFK, LGA, and EWR) in 2013. As you can probably imagine, this isn't a small dataset.

```{r}
# install.packages('nycflights13')
library(nycflights13)
data(flights)

names(flights)
```

## Data linkage

Suppose we want to know more about the airline (`carrier`). In the data set, each carrier is stored using a two-letter code.

```{r}
table(flights$carrier)
```

- Why use a two-letter code instead of the airline name?
- Can we _link_ the airline names to the letter codes?

Use a common variable, called a _key_, to link the data.

## `inner_join()`

```{r}
data(airlines)

flights_carrier <- flights %>%
  inner_join(airlines, by=c('carrier'='carrier'))
```

Did it work?

```{r}
names(flights_carrier)
glimpse(flights_carrier)
```

## Exercises: Departure delays on United

In teams of 2-3, complete the following exercises:

1. Create a new data set, `flights2` that contains the carrier name, year, month, day, departure delay, arrival delay, origin airport, destination airport, and flight number.

```{r}

```

2. How many unique flight routes does United Airlines run that depart the New York area?

```{r}

```

3. How many unique destinations does United Airlines serve from the New York Area?

```{r}

```

4. How many unique flight routes does United Airlines run from each of the three area airports?

```{r}

```

5. What is the average departure delay of a United Airlines flight leaving any New York area airport?

```{r}

```

6. What is the average departure delay of a United Airlines flight leaving JFK? LGA? EWR?

```{r}

```


## Example: Relative age effect

The "Relative Age Effect" is an attempt to explain anomalies in the distribution of birth month among athletes. Briefly, the idea is that children born just after the age cut-off for participation will be as much as 11 months older than their fellow athletes, which is enough of a disparity to give them an advantage. This advantage gets cmpounded over the years, resulting in notably more professional athletes born in these months. 

Is there evidence of the relative age effect in professional baseball?

This database contains pitching, hitting, and fielding statistics for Major League Baseball from 1871 through 2016. It includes data from the two current leagues (American and National), the four other "major" leagues (American Association, Union Association, Players League, and Federal League), and the National Association of 1871-1875.

```{r}
#install.packages('Lahman')
library(Lahman)
```

Let's start with pitchers.

```{r}
data(Pitching)
names(Pitching)
```

Check the documentation for the `Lahman` package. How could we get the birthdays of professional pitchers?

```{r}
data(Master)
names(Master)
pitcher_bday <- Pitching %>%
  inner_join(Master, by=c('playerID'='playerID'))
```

Let's simplify this data set to include only player ID and birth month.

```{r}
pitcher_bday <- pitcher_bday %>% select(c('playerID', 'birthMonth'))
pitcher_bday %>% group_by(birthMonth) %>% 
  summarize(N=n())
```

__Problem__: Are players unique?

```{r}
pitcher_bday %>% arrange(playerID)
```

Run this code in your notes and see what happens. (Comment it out before you compile your document - trust me.)

```{r}
pitcher_bday %>% distinct() %>% 
  group_by(birthMonth) %>% 
  summarize(N=n())
```

```{r}
unique_bday <- pitcher_bday %>% 
  distinct() %>% 
  group_by(birthMonth) %>% 
  summarize(N=n())
```

```{r, warning=FALSE}
ggplot(pitcher_bday, aes(x=birthMonth)) + 
  geom_bar(aes(fill=birthMonth))
```

```{r, eval=FALSE, warning=FALSE}
ggplot(pitcher_bday, aes(x=as.factor(birthMonth))) +
  geom_bar(aes(fill=as.factor(birthMonth))) +
  labs(x='Birth Month') +
  theme(legend.position="none")
```

```{r, echo=FALSE, warning=FALSE}
ggplot(pitcher_bday, aes(x=as.factor(birthMonth))) +
  geom_bar(aes(fill=as.factor(birthMonth))) +
  labs(x='Birth Month') +
  theme(legend.position="none")
```

Is there evidence of a relative age effect for batters? Hint: Use the `Lahman` package documentation to get started.

https://cran.r-project.org/web/packages/Lahman/Lahman.pdf

Use the `viridis` online documentation to choose a different color scale.

https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html
