---
title: 'Section 2: Data Visualization'
author: 'MTH 365: Introduction to Data Science'
date: "`r Sys.Date()`"
output: 
  ioslides_presentation:
    widescreen: true
---

<style>
h2 { 
 color: #0054a6;		
}
h3 { 
 color: #019cdb;	
}
</style>

## Apologies for cross-teaching

Sorry MTH 361 students.

## Recommended Reading and Packages

- _Modern Data Science with R_ Ch. 2: Data Visualization
- _Modern Data Science with R_ Ch. 3: A Grammar for Graphics

```{r, warning=FALSE, message=FALSE}
#install.packages(dslabs)
library(dslabs)

#install.packages(tidyverse)
library(tidyverse)
```

## `ggplot2`

`ggplot2` is an R package for "declaratively creating graphics"

- https://ggplot2.tidyverse.org/reference/

\vfill

\textcolor{CosmoBlue}{Basic Format of a Plot}

`data and aesthetics + plot type + options`

## Hate crimes and income inequality

<div class="red2">__Example__:</div>

A FiveThirtyEight article published in 2017 claimed that higher rates of hate crimes were tied to greater income inequality.

https://fivethirtyeight.com/features/higher-rates-of-hate-crimes-are-tied-to-income-inequality/

- FiveThirtyEight publishes their data sets - let's investigate for ourselves.

## Hate crimes and income inequality

```{r, echo=-1}
hate_crimes <- read.csv("~/OneDrive - Creighton University/Fall 2019 Courses/MTH 365 - Intro to Data Science/Data/hate_crimes.csv")
glimpse(hate_crimes)
```

## A simple plot?

```{r, warning=FALSE}
ggplot(hate_crimes, aes(x=median_income, y=hate_crimes_fbi)) +
  geom_point()
```

## A simple plot

```{r, warning=FALSE}
ggplot(hate_crimes, aes(x=median_income, y=hate_crimes_fbi))
```

## A better plot 

```{r, warning=FALSE}
ggplot(hate_crimes, aes(x=median_income, y=hate_crimes_fbi)) +
  geom_text(aes(label=state))
```

## An even better plot 

```{r, warning=FALSE}
ggplot(hate_crimes, aes(x=median_income, y=hate_crimes_fbi)) + 
  geom_text(aes(label=state, color=region))
```

## Measuring income inequality 

```{r, warning=FALSE}
ggplot(hate_crimes, aes(x=gini_index, y=hate_crimes_fbi)) + 
  geom_point(aes(color=region))
```

## Does source matter? 

```{r, warning=FALSE}
ggplot(hate_crimes, aes(x=gini_index, y=hate_crimes_splc)) + 
  geom_point(aes(color=region))
```

## Adding labels 

```{r, warning=FALSE}
ggplot(hate_crimes, aes(x=gini_index, y=hate_crimes_fbi)) + 
  geom_point(aes(color=region)) + 
  labs(x='Gini Index', y='Hate Crimes (FBI, Sept. 2016)')
```

## Point size 

```{r, warning=FALSE}
ggplot(hate_crimes, aes(x=gini_index, y=hate_crimes_fbi)) + 
  geom_point(aes(color=region, size=unemployment)) + 
  labs(x='Gini Index', y='Hate Crimes (FBI, Sept. 2016)')
```

## Adding "facets" 

```{r, warning=FALSE}
ggplot(hate_crimes, aes(x=gini_index, y=hate_crimes_fbi)) + 
  geom_point(aes(color=region, size=unemployment)) + 
  labs(x='Gini Index', y='Hate Crimes (FBI, Sept. 2016)')
  facet_wrap(~region, nrow=1)
```

## Hate crimes and income inequality

<div class="green2">__Exploration__:</div>

In teams of two or three, take a few minutes to create some new visualizations using this data. What else can you learn about hate crimes and income inequality?

```{r}
names(hate_crimes)
```

## Weather patterns

<div class="red2">__Example__:</div>

The data set `Weather` contains data on weather-related variables for several world cities.

```{r}
#install.packages(mosaicData)
library(mosaicData)
data(Weather)
```

## Weather patterns

```{r}
glimpse(Weather)
```

## Weather patterns

```{r}
ggplot(Weather, aes(y=avg_temp)) + geom_boxplot()
```

## Weather patterns

```{r}
ggplot(Weather, aes(y=avg_temp, group=city)) + geom_boxplot(aes(color=city))
```

## Weather patterns

```{r}
ggplot(Weather, aes(x=date, y=avg_temp))+geom_point()
```


## Weather patterns

```{r}
ggplot(Weather, aes(x=date, y=avg_temp)) + 
  geom_point(aes(color=city))
```

## Weather patterns in Chicago

```{r}
Chicago <- Weather %>% filter(city=='Chicago')
ggplot(Chicago, aes(x=date, y=avg_temp)) + 
  geom_point(color='blue')
```

## Weather patterns in Chicago

```{r}
ggplot(Chicago, aes(x=date, y=avg_temp)) + 
  geom_point(aes(color=events))
```

## Weather patterns in Chicago

```{r}
ggplot(Chicago, aes(x=date, y=avg_temp)) + 
  geom_smooth()
```

## Weather patterns in Chicago

```{r}
ggplot(Chicago, aes(x=date, y=avg_temp)) + 
  geom_smooth()+geom_point()
```

## Weather patterns in Chicago

```{r}
ggplot(Chicago, aes(x=low_temp, y=high_temp)) + 
  geom_point(aes(color=month))
```

## Weather patterns in Chicago

```{r}
ggplot(Chicago, aes(x=low_temp, y=high_temp)) + 
  geom_point(aes(color=as.factor(month)))
```

## Weather patterns in Chicago

```{r}
ggplot(Chicago, aes(x=avg_temp)) + 
  geom_histogram()
```

## Weather patterns in Chicago

```{r}
ggplot(Chicago, aes(x=avg_temp)) + 
  geom_density(fill='lightblue')
```

## Weather patterns in Chicago

```{r}
ggplot(Chicago, aes(x=avg_temp)) + 
  geom_density(fill='lightblue', alpha=0.5)
```

## Weather patterns in Chicago

```{r}
ggplot(Chicago, aes(x=avg_temp, group=month)) +
  geom_density(aes(color=as.factor(month), 
                   fill=as.factor(month)), alpha=0.5)
```

## Weather patterns in Chicago

```{r}
ggplot(Chicago, aes(x=avg_temp, group=month)) +
  geom_density(aes(color=as.factor(month),                                                    fill=as.factor(month)), alpha=0.5) +
  facet_wrap(~month, nrow=3)
```

## Weather patterns in Chicago

```{r}
ggplot(Chicago, aes(x=avg_vis, group=month)) +
  geom_density(aes(color=as.factor(month), 
                   fill=as.factor(month)), alpha=0.5) +
  facet_wrap(~month, nrow=3)
```

## Weather events

```{r}
ggplot(Weather, aes(x=events)) + 
  geom_bar()
```

## Weather events

```{r}
ggplot(Weather, aes(x=events)) + 
  geom_bar() + 
  coord_flip()
```

## Weather events in Chicago

```{r}
ggplot(Chicago, aes(x=events)) + 
  geom_bar() + 
  coord_flip() +
  facet_wrap(~year, nrow=2)
```
