---
title: "Section 2: Data Visualization (Part 1)"
author: 'MTH 365: Introduction to Data Science'
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

## Recommended Reading

- _Modern Data Science with R_ Ch. 2: Data Visualization
- _Modern Data Science with R_ Ch. 3: A Grammar for Graphics

```{r, message=FALSE}
#install.packages(dslabs)
library(dslabs)
```

## `ggplot2`

`ggplot2` is an R package for "declaratively creating graphics"

- https://ggplot2.tidyverse.org/reference/

__Basic Format of a Plot__:

`data and aesthetics + plot type + options`

```{r, message=FALSE}
#install.packages(tidyverse)
library(tidyverse)
library(ggplot2)
```

- These warnings may appear if your version of R is out of date with the package version. Most of the time, this isn't a problem! If you want to suppress it, add `warning=FALSE` as a chunk option.

## Example: Hate crimes and income inequality

A FiveThirtyEight article published in 2017 claimed that higher rates of hate crimes were tied to greater income inequality.

https://fivethirtyeight.com/features/higher-rates-of-hate-crimes-are-tied-to-income-inequality/

- FiveThirtyEight publishes their data sets - let's investigate for ourselves.
- Data sets are posted on BlueLine. Download this data set, and save it to your computer. 

Follow these steps to read the data into RStudio:

1. In the Environment tab, click "Import Dataset". Since this is a CSV document, you want to import a text file.
2. Navigate to your CSV data set. Make sure that the first row contains column names. 
3. Import the data.

You'll see a line of code pop up in your R console, it should start with `read.csv()`. Copy this code and replace the provided statement (`hate_crimes <- read.csv("~/OneDrive - Creighton University/Fall 2019 Courses/MTH 365 - Intro to Data Science/Data/hate_crimes.csv")`) with the one you just generated.

```{r}
hate_crimes <- read.csv("~/OneDrive - Creighton University/2019/Spring 2019 Courses/MTH 365 - Intro to Data Science/Data/hate_crimes.csv")
glimpse(hate_crimes)
```

## A simple plot?

```{r}
ggplot(hate_crimes, aes(x=median_income, y=hate_crimes_fbi)) +
  geom_point(aes(col=state))
```

```{r}
ggplot(hate_crimes, aes(x=median_income, y=hate_crimes_fbi))
```

```{r}
ggplot(hate_crimes, aes(x=median_income, y=hate_crimes_fbi)) + 
  geom_text(aes(label=state))
```

```{r}
ggplot(hate_crimes, aes(x=median_income, y=hate_crimes_fbi)) +
  geom_text(aes(label=state, color=region))

ggplot(hate_crimes, aes(x=median_income, y=hate_crimes_fbi)) +
  geom_text(aes(label=state), nudge_x=1000, nudge_y=1) + 
  geom_point(aes(col=region, pch=region), cex=2)
```

## Measuring income inequality

```{r}
hate_crimes %>% filter(state != 'District of Columbia') %>%
  ggplot(aes(x=gini_index, y=hate_crimes_fbi)) +
  geom_point(aes(color=region, pch=region), cex=5)
```

Does source matter?

```{r}
ggplot(hate_crimes, aes(x=gini_index, y=hate_crimes_splc)) +
  geom_point(aes(color=region))
```

## Adding plot options

```{r}
ggplot(hate_crimes, aes(x=gini_index, y=hate_crimes_fbi)) +
  geom_point(aes(color=region)) + 
  labs(x='Gini Index', y='Hate Crimes (FBI, Sept. 2016)')
```

```{r}
ggplot(hate_crimes, aes(x=gini_index, y=hate_crimes_fbi)) +
  geom_point(aes(color=region, size=unemployment)) + 
  labs(x='Gini Index', y='Hate Crimes (FBI, Sept. 2016)')
```

```{r}
ggplot(hate_crimes, aes(x=gini_index, y=hate_crimes_fbi)) +
  geom_point(aes(color=region, size=sqrt(unemployment))) + 
  labs(x='Gini Index', y='Hate Crimes (FBI, Sept. 2016)') + 
  facet_wrap(~region, ncol=3)
```

<div class="green2">__Exploration__:</div>

In teams of two or three, take a few minutes to create some new visualizations using this data. What else can you learn about hate crimes and income inequality?

```{r}
# Show and tell plots
hate_crimes %>% 
  ggplot(aes(x=hate_crimes_fbi, y=hate_crimes_splc)) + 
  geom_point(aes(col=unemployment))




```

## Weather patterns

The data set `Weather` contains data on weather-related variables for several world cities.

```{r}
#install.packages("mosaicData")
library(mosaicData)
data(Weather)
glimpse(Weather)
```

```{r}
ggplot(Weather, aes(y=avg_temp)) + 
  geom_boxplot()
```

```{r}
ggplot(Weather, aes(y=avg_temp, group=city)) + 
  geom_boxplot(aes(color=city))
```

```{r}
ggplot(Weather, aes(x=date, y=avg_temp)) +
  geom_point()
```

```{r}
ggplot(Weather, aes(x=date, y=avg_temp)) +
  geom_point(aes(color=city))
```

### Weather patterns in Chicago

```{r}
Chicago <- Weather %>% filter(city=='Chicago')
ggplot(Chicago, aes(x=date, y=avg_temp)) +
  geom_point(col='blue')
```

```{r}
ggplot(Chicago, aes(x=date, y=avg_temp)) +
  geom_point(aes(color=events))
```

```{r}
ggplot(Chicago, aes(x=date, y=avg_temp)) +
  geom_smooth()
```

```{r}
ggplot(Chicago, aes(x=date, y=avg_temp)) +
  geom_smooth() +
  geom_point()
```

```{r}
glimpse(Chicago)

ggplot(Chicago, aes(x=low_temp, y=high_temp)) +
  geom_point(aes(color=month)) + 
  
```

```{r}
ggplot(Chicago, aes(x=low_temp, y=high_temp)) +
  geom_point(aes(color=as.factor(month)))
```

```{r}
ggplot(Chicago, aes(x=avg_temp)) + 
  geom_histogram()
```

```{r}
ggplot(Chicago, aes(x=avg_temp)) + 
  geom_density(fill='lightblue')
```

```{r}
ggplot(Chicago, aes(x=avg_temp)) + 
  geom_density(fill='lightblue', alpha=0.5)
```

```{r}
ggplot(Chicago, aes(x=avg_temp, group=month)) + 
  geom_density(aes(color=as.factor(month), fill=as.factor(month)), alpha=0.5)
```

```{r}
ggplot(Chicago, aes(x=avg_temp, group=month)) +
  geom_density(aes(color=as.factor(month), fill=as.factor(month)), alpha=0.5) + 
  facet_wrap(~month, nrow=3)
```

```{r}
ggplot(Chicago, aes(x=avg_vis, group=month)) + 
  geom_density(aes(color=as.factor(month), fill=as.factor(month)), alpha=0.5) +
  facet_wrap(~month, nrow=3)
```

## Weather events

```{r}
ggplot(Weather, aes(x=events)) +
  geom_bar()
```

```{r}
ggplot(Weather, aes(x=events)) +
  geom_bar() +
  coord_flip()
```

### Weather events in Chicago

```{r}
ggplot(Chicago, aes(x=events)) +
  geom_bar() +
  coord_flip() +
  facet_wrap(~year, nrow=2)
```