---
title: 'Section 13: Text Mining'
author: 'MTH 365: Introduction to Data Science'
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  xaringan::moon_reader:
    css: ["style.css", 'hygge']
    lib_dir: libs
    nature:
      highlightStyle: magula
      highlightLines: true
      countIncrementalSlides: false
---

## Recommended Reading

- _Tidy Text Mining with R_: Julia Silge and David Robinson (https://www.tidytextmining.com/)

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(mdsr)
library(RColorBrewer)
library(tidytext)
```

Like the last section, this document takes a _very_ long time to knit. Don't do it until the end of class.

---

## Jeopardy!

__Example__: Reddit user u/PandaPython1 posted web scraped data from the Jeopardy Archive (http://www.jarchive.com) containing the questions, categories, answers, clue locations, and earnings from all Jeopardy shows through the 2017 season. The data is saved as `Jeopardy2017.csv`.

```{r}
Jeopardy2017 <- read.csv("~/OneDrive - Creighton University/Fall 2019 Courses/MTH 365 - Intro to Data Science/Data/Jeopardy2017.csv")
names(Jeopardy2017)
```

---

## Jeopardy!

If you're a future Jeopardy contestant, you might want to limit your studying:

1. What categories are most common?
2. What types of answers are most common?

```{r}
Jeopardy2017 %>% summarize(n=n_distinct(categories))
```

---

## Most common categories and answers

Options: 

1. Review all 38,823 categories for classification manually
2. Use text mining!

.full-width[.content-box-yellow[__Text mining__: the process of extracting high-quality information from text. ]]

- High-quality information is typically derived through the devising of patterns and trends through means such as statistical pattern learning.

---

## Most common categories and answers

```{r, eval=FALSE}
category_df <- tibble(question=1:nrow(Jeopardy2017),
          category=as.character(Jeopardy2017$categories))
tidy_category <- category_df %>% 
  unnest_tokens(word, category) 
tidy_category %>% group_by(word) %>% 
  summarize(n=n())
```

---

## Most common categories and answers

```{r, echo=FALSE}
category_df <- tibble(question=1:nrow(Jeopardy2017),
          category=as.character(Jeopardy2017$categories))
tidy_category <- category_df %>% 
  unnest_tokens(word, category) 
tidy_category %>% group_by(word) %>% 
  summarize(n=n())
```


---

## Most common categories and answers

```{r, eval=FALSE}
tidy_category %>% group_by(word) %>% 
  summarize(n=n()) %>% 
  arrange(desc(n))
```

---

## Most common categories and answers

```{r, echo=FALSE}
tidy_category %>% group_by(word) %>% 
  summarize(n=n()) %>% 
  arrange(desc(n))
```

---

## Stop words

.full-width[.content-box-yellow[__Stop words__: words that are not useful for an analysis, typically extremely common words such as "the", "of", "to", and so forth in English]]

```{r, eval=FALSE}
data(stop_words)
tidy_category <- tidy_category %>% anti_join(stop_words)
```

- What do you think `anti_join()` does?

---

```{r}
head(tidy_category)
data(stop_words)
tidy_category <- tidy_category %>% anti_join(stop_words)
head(tidy_category)
```


---

## Stop words

```{r, eval=FALSE}
tidy_category %>% group_by(word) %>% 
  summarize(n=n()) %>% 
  arrange(desc(n))
```

---

## Stop words

```{r, echo=FALSE}
tidy_category %>% group_by(word) %>% 
  summarize(n=n()) %>% 
  arrange(desc(n))
```

---

## Most common categories

```{r, eval=FALSE}
top_categories <- tidy_category %>% group_by(word) %>% 
  summarize(n=n()) %>% 
  filter(n>=2000)
top_categories %>% ggplot(aes(x=word, y=n)) + 
  geom_col() + coord_flip()
```

---

## Most common categories

```{r, echo=FALSE}
top_categories <- tidy_category %>% group_by(word) %>% 
  summarize(n=n()) %>% 
  filter(n>=2000)
top_categories %>% ggplot(aes(x=word, y=n)) + 
  geom_col() + coord_flip()
```

---

## Category bigrams

```{r, eval=FALSE}
category_bigrams <- category_df %>%
  unnest_tokens(bigram, category, token = "ngrams", n = 2)
category_bigrams %>% group_by(bigram) %>% 
  summarize(n=n()) %>% arrange(desc(n))
```

---

## Category bigrams

```{r, echo=FALSE}
category_bigrams <- category_df %>%
  unnest_tokens(bigram, category, token = "ngrams", n = 2)
category_bigrams %>% group_by(bigram) %>% 
  summarize(n=n()) %>% arrange(desc(n))
```

---

## Category bigrams

```{r}
bigrams_separated <- category_bigrams %>%
  separate(bigram, c("word1", "word2"), sep = " ")

bigrams_filtered <- bigrams_separated %>%
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word)

# new bigram counts:
bigram_counts <- bigrams_filtered %>% 
  count(word1, word2, sort = TRUE) %>% 
  filter(word1 != 'NA') %>% 
  filter(word2 != 'NA')
```

---

## Category bigrams

```{r}
bigram_counts
```

---

## Category bigrams

```{r, echo=FALSE}
top_bigrams <- bigram_counts %>% filter(n>=300) %>% mutate(bigram = paste(word1, ' ', word2))
top_bigrams %>% ggplot(aes(x=reorder(bigram, n), y=n)) + geom_col() + coord_flip()
```

---

## Most common questions

```{r, echo=FALSE, message=FALSE}
#Jeopardy2017 %>% summarize(n=n_distinct(categories))

question_df <- tibble(number=1:nrow(Jeopardy2017),
                      question=as.character(Jeopardy2017$question))
tidy_question <- question_df %>% unnest_tokens(word, question) 

tidy_question <- tidy_question %>% anti_join(stop_words)

top_questions <- tidy_question %>% group_by(word) %>% summarize(n=n()) %>% filter(n>=3000)
top_questions %>% ggplot(aes(x=reorder(word, n), y=n)) + geom_col() + coord_flip()
```

---

## Question bigrams

```{r, echo=FALSE}
question_bigrams <- question_df %>%
  unnest_tokens(bigram, question, token = "ngrams", n = 2)

bigrams_separated <- question_bigrams %>%
  separate(bigram, c("word1", "word2"), sep = " ")

bigrams_filtered <- bigrams_separated %>%
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word)

# new bigram counts:
bigram_counts <- bigrams_filtered %>% 
  count(word1, word2, sort = TRUE) %>% 
  filter(word1 != 'NA') %>% 
  filter(word2 != 'NA')

top_bigrams <- bigram_counts %>% filter(n>=300) %>% mutate(bigram = paste(word1, ' ', word2))
top_bigrams %>% ggplot(aes(x=reorder(bigram, n), y=n)) + geom_col() + coord_flip()
```

---

## Most common answers

__Example__: On your own or with a partner, use text mining to investigate the most common Jeopardy answers. If you're planning an appearance on Jeopardy, what should you brush up on?

---

## Most common answers

```{r, echo=FALSE, message=FALSE}
answer_df <- tibble(number=1:nrow(Jeopardy2017),
                      answer=as.character(Jeopardy2017$answer))
tidy_answer <- answer_df %>% unnest_tokens(word, answer) 

tidy_answer <- tidy_answer %>% anti_join(stop_words)
top_answers <- tidy_answer %>% group_by(word) %>% summarize(n=n()) %>% filter(n>=2000)

answer_bigrams <- answer_df %>%
  unnest_tokens(bigram, answer, token = "ngrams", n = 2)

bigrams_separated <- answer_bigrams %>%
  separate(bigram, c("word1", "word2"), sep = " ")

bigrams_filtered <- bigrams_separated %>%
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word)

# new bigram counts:
bigram_counts <- bigrams_filtered %>% 
  count(word1, word2, sort = TRUE) %>% 
  filter(word1 != 'NA') %>% 
  filter(word2 != 'NA')

top_bigrams <- bigram_counts %>% filter(n>=75) %>% mutate(bigram = paste(word1, ' ', word2))
top_bigrams %>% filter(n > 50) %>% ggplot(aes(x=reorder(bigram, n), y=n)) + geom_col() + coord_flip()
```

---

## Exploring Harry Potter

__Example__: The R library `harrypotter` contains the text of all seven Harry Potter books.

```{r, eval=FALSE}
#devtools::install_github("bradleyboehmke/harrypotter")
library(harrypotter)
philosophers_stone[1]
```

```{r}
library(harrypotter)
```

---

## Sentiment analysis

.full-width[.content-box-yellow[__Sentiment analysis__: the process of classifying a given text into a certain category, such as emotional states]]

- Examples: anger, anticipation, constraining, disgust, fear, joy, litigious, negative, positive, sadness, superfluous, surprise, trust, uncertainty

---

## Sentiment analysis

```{r}
sentiments
```

---

## Sentiment analysis

UPDATE FOR NEW SENTIMENTS

Lexicon choices:

- `nrc`: categorizes words into categories of emotions (positive, negative, anger, anticipation, disgust, fear, joy, sadness, surprise, trust)
- `bing`: categorizes words into positive and negative categories
- `AFINN`: assigns words a score between -5 and +5, with negative scores indicating negative sentiment (-5 and -4 words are NSFW)

---

## Sentiment analysis: `nrc`

```{r}
sentiments %>% filter(lexicon=='nrc')
```

---

## Sentiment analysis: `bing`

```{r}
sentiments %>% filter(lexicon=='bing')
```

---

## Sentiment analysis: `AFINN`

```{r}
sentiments %>% filter(lexicon=='AFINN')
```

---

## Sentiments in Harry Potter

```{r}
titles <- c("Philosopher's Stone", 
            "Chamber of Secrets", 
            "Prisoner of Azkaban",
            "Goblet of Fire", 
            "Order of the Phoenix", 
            "Half-Blood Prince",
            "Deathly Hallows")

books <- list(philosophers_stone, 
              chamber_of_secrets, 
              prisoner_of_azkaban,
              goblet_of_fire, 
              order_of_the_phoenix, 
              half_blood_prince,
              deathly_hallows)
  
series <- tibble()
```

---

## Sentiments in Harry Potter

```{r}
for(i in seq_along(titles)) {
        
        clean <- tibble(chapter = seq_along(books[[i]]),
                        text = books[[i]]) %>%
             unnest_tokens(word, text) %>%
             mutate(book = titles[i]) %>%
             select(book, everything())

        series <- rbind(series, clean)
}

series$book <- factor(series$book, levels = rev(titles))
```

---

## Sentiments in Harry Potter

```{r, eval=FALSE}
series %>%
        right_join(get_sentiments("nrc")) %>%
        filter(!is.na(sentiment)) %>%
        count(sentiment, sort = TRUE)
```

---

## Sentiments in Harry Potter

```{r, echo=FALSE, message=FALSE}
series %>%
        right_join(get_sentiments("nrc")) %>%
        filter(!is.na(sentiment)) %>%
        count(sentiment, sort = TRUE)
```

---

## Sentiments in Harry Potter

```{r, message=FALSE}
sentiment_by_book <- series %>% group_by(book) %>% 
  mutate(word_count = 1:n(), index = word_count%/%500+1) %>%
  inner_join(get_sentiments("bing")) %>%
  count(book, index = index , sentiment) %>%
  ungroup() %>%
  spread(sentiment, n, fill = 0) %>%
  mutate(sentiment = positive - negative, 
         book = factor(book, levels = titles))
```

---

## Sentiments in Harry Potter

```{r}
sentiment_by_book
```

---

## Sentiments in Harry Potter

```{r, eval=FALSE}
sentiment_by_book %>% ggplot(aes(index, sentiment, fill = book)) +
  geom_bar(alpha = 0.5, stat = "identity", show.legend = FALSE) +
  facet_wrap(~ book, ncol = 2, scales = "free_x")
```

---

## 

```{r, echo=FALSE, fig.height=9, fig.width=10}
sentiment_by_book %>% ggplot(aes(index, sentiment, fill = book)) +
  geom_bar(alpha = 0.5, stat = "identity", show.legend = FALSE) +
  facet_wrap(~ book, ncol = 2, scales = "free_x")
```

---

## "The Philosopher's Stone"

```{r, echo=FALSE, fig.width=10, fig.height=6}
sentiment_by_book %>% filter(book=="Philosopher's Stone") %>% ggplot(aes(index, sentiment, fill = book)) +
  geom_bar(alpha = 0.5, stat = "identity", show.legend = FALSE) + 
  labs(title="The Philosopher's Stone")
```

---

## "The Deathly Hallows"

```{r, echo=FALSE, fig.width=10, fig.height=6}
sentiment_by_book %>% filter(book=="Deathly Hallows") %>% ggplot(aes(index, sentiment, fill = book)) +
  geom_bar(alpha = 0.5, stat = "identity", show.legend = FALSE) + 
  labs(title="The Deathly Hollows")
```

---

## Sentiments in Harry Potter

```{r}
sentiment_by_book %>% group_by(book) %>%
  summarize(mean=mean(sentiment),
            sd=sd(sentiment),
            min=min(sentiment),
            max=max(sentiment))
```

---

## Sentiments in Harry Potter: `AFINN`

```{r, eval=FALSE}
sentiment_afinn <- series %>% group_by(book) %>% 
  mutate(word_count = 1:n(), index = word_count%/%500+1) %>%
  inner_join(get_sentiments("afinn")) %>%
  count(book, index = index , score) %>%
  ungroup() %>%
  spread(score, n, fill = 0) %>%
  mutate(score = -5*`-5`+-4*`-4`+-3*`-3`+
           -2*`-2`+-1*`-1`+1*`1`+2*`2`+
           3*`3`+4*`4`+5*`5`, 
         book = factor(book, levels = titles))
sentiment_afinn
```

---

## Sentiments in Harry Potter: `AFINN`

```{r, echo=FALSE, message=FALSE}
sentiment_afinn <- series %>% group_by(book) %>% 
  mutate(word_count = 1:n(), index = word_count%/%500+1) %>%
  inner_join(get_sentiments("afinn")) %>%
  count(book, index = index , score) %>%
  ungroup() %>%
  spread(score, n, fill = 0) %>%
  mutate(score = -5*`-5`+-4*`-4`+-3*`-3`+
           -2*`-2`+-1*`-1`+1*`1`+2*`2`+
           3*`3`+4*`4`+5*`5`, 
         book = factor(book, levels = titles))
sentiment_afinn
```

---

```{r, echo=FALSE, fig.width=10, fig.height=9}
sentiment_afinn %>% ggplot(aes(index, score, fill = book)) +
  geom_bar(alpha = 0.5, stat = "identity", show.legend = FALSE) +
  facet_wrap(~ book, ncol = 2, scales = "free_x")
```

---

## Sentiments in Harry Potter: `AFINN`

```{r}
sentiment_afinn %>% group_by(book) %>%
  summarize(mean=mean(score),
            sd=sd(score),
            min=min(score),
            max=max(score))
```