---
title: 'Section 3: Data Wrangling (Part 1)'
author: 'MTH 365: Introduction to Data Science'
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

## Recommended Reading

- _Modern Data Science with R_ Ch. 4: Data Wrangling
- _Modern Data Science with R_ Ch. 5: Tidy Data and Iteration
- Wickham, Hadley. (2014). "Tidy Data". _Journal of Statistical Software_ 59(10). Available on BlueLine.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

## Data structure and semantics

- Most statistical datasets are tables made up of _rows_ and _columns_

A dataset is a collection of _values_: these can be _numbers_ (quantiative) or character _strings_ (qualitative)

- Every value belongs to a variable and an observation

__Variable__: 

__Observation__: 

## Tidy data

"Tidy" data is a standard way of mapping the meaning of a dataset to its structure. 

1. Each variable forms a column.
2. Each observation forms a row.
3. Each type of observational unit forms a table.

Any other arrangement of the data is called "messy".

Example: Tables 1 and 2 rearranged in tidy format.

person|treatment|result
---|---|---
John Smith|a|-
Jane Doe|a|16
Mary Johnson|a|3
John Smith|b|2
Jane Doe|b|11
Mary Johnson|b|1

Real datasets can, and often do, violate the three principles of tidy data in almost every way imaginable! 

## Example: Gapminder

Gapminder is an independent Swedish foundation with no political, religious or economic affiliations. __Gapminder is a fact tank, not a think tank.__ Gapminder fights devastating misconceptions about global development. Gapminder produces free teaching resources making the world understandable based on reliable statistics. Gapminder promotes a fact-based worldview everyone can understand.  Gapminder collaborates with universities, UN, public agencies and non-governmental organizations.

## Loading the Gapminder data set

```{r}
library(dslabs)
data(gapminder)
glimpse(gapminder)
```

## Exploring the Gapminder data set

```{r, warning=FALSE}
ggplot(gapminder, aes(x=fertility, y=infant_mortality)) +
  geom_point()
```

```{r, warning=FALSE}
ggplot(gapminder, aes(x=fertility, y=infant_mortality)) +
  geom_point(aes(col=country))
```

```{r, warning=FALSE}
ggplot(gapminder, aes(x=fertility, y=infant_mortality)) +
  geom_point(aes(col=region))
```

```{r, warning=FALSE}
ggplot(gapminder, aes(x=fertility, y=infant_mortality)) +
  geom_point(aes(col=region)) + 
  facet_wrap(aes(year))
```

## Data wrangling with `dplyr`

There are five major `dplyr` functions for wrangling data:

- `select()`: take a subset of the columns (variables/features)
- `filter()`: take a subset of the rows (observations)
- `mutate()`: add or modify existing columns
- `arrange()`: sort the rows
- `summarize()`: aggregate the data across rows (grouping)

## `dplyr` rules for functions


## Piping v. layering

Pipe or layer?

## `select()`

- `select()`: 


```{r}
gapminder_short <- gapminder %>% select(gdp, region)
glimpse(gapminder_short)
```

## `filter()`

- `filter()`: 

```{r}
gapminder2000 <- gapminder %>% 
  filter(year==2000)
glimpse(gapminder2000)

# Extract NAs?
gapminderNA <- gapminder %>%
  filter(is.na(gdp))
glimpse(gapminderNA)
```

```{r, warning=FALSE}
ggplot(gapminder2000, aes(x=fertility, y=infant_mortality)) +
  geom_point(aes(col=region))
```

```{r, eval=FALSE, warning=FALSE}
gapminder %>% filter(year==2000) %>%
  filter(region %in% c('Central Asia', 
                       'Eastern Asia', 
                       'Polynesia', 
                       'South-Eastern Asia', 
                       'Southern Asia', 
                       'Western Asia')) %>%
  ggplot(aes(x=fertility, y=infant_mortality)) +
  geom_point(aes(col=region))
```

```{r, warning=FALSE}
gapminder %>% filter(year==2000) %>%
  filter(region %in% c('Central Asia', 'Eastern Asia', 
                       'Polynesia', 'South-Eastern Asia', 
                       'Southern Asia', 'Western Asia')) %>%
  ggplot(aes(x=fertility, y=infant_mortality)) +
  geom_point(aes(col=region))
```

## (Smooth) Operators

- `%>%`: 
- `%in%`:
- `==`: 

## More Operators

Operator|Definition                   
---|---
`<`| less than                    
`<=`|	less than or equal to
`>`| greater than
`>=`|	greater than or equal to 
`==`|	exactly equal to 
`!=`|	not equal to 
`x & y`| `x` AND `y`                  
`x %in% y`| test if `x` is in `y`   


## Fertility and infant mortality in the US

Example: In your notes, modify your plot for another country. Describe how fertility and infant mortality are changing over time.

```{r, warning=FALSE}
gapminder %>% 
  filter(country %in% c('Italy', 'United States', 'China')) %>%
  ggplot(aes(x=fertility, y=infant_mortality)) +
  geom_point(aes(col=year, pch=country))

gapminder %>% filter(country=='China') %>% filter(infant_mortality >= 1)
```

## `mutate()`

- `mutate()` lets us create new variables based on manipulations of the old variables

Example: We'd like to calculate the gross domestic product per capita. Here are the variables in our data - write an expression to do this calculation.

```{r}
names(gapminder)
```

```{r}
gapminder <- gapminder %>% 
  mutate(GDP_pc = gdp/population,
         mortality_fertility = infant_mortality/fertility)
glimpse(gapminder)
```

Example: How do we find out which countries have the highest GDP per capita?

```{r, warning=FALSE, message=FALSE}
ggplot(gapminder, aes(x=GDP_pc))+geom_density()
```

Example: How do we find out which countries have the highest GDP per capita?

```{r}
high_GDP <- gapminder %>% filter(GDP_pc>50000)
```

```{r}
high_GDP
```

## `arrange()`

Right now, the data is sorted by country, then year. We could use the `arrange()` command to resort in terms of another variable.

```{r, eval=-2}
GDP_only <- gapminder %>% select(country, year, GDP_pc, continent)
GDP_only %>% arrange(desc(GDP_pc))
```

```{r, eval=FALSE}
GDP_only %>% arrange(continent, desc(GDP_pc))
```

## `summarize()`

If we want to compare summary statistics, we might use `summarize()`.

```{r}
GDP_only %>% summarize(avg_GDP_pc=mean(GDP_pc), 
                       sd_GDP_pc=sd(GDP_pc), 
                       min_GDP_pc=min(GDP_pc), 
                       max_GDP_pc=max(GDP_pc))
```

```{r}
GDP_only %>% filter(GDP_pc != "NA") %>%
  summarize(avg_GDP_pc=mean(GDP_pc), 
            sd_GDP_pc=sd(GDP_pc), 
            min_GDP_pc=min(GDP_pc), 
            max_GDP_pc=max(GDP_pc),
            N=n())
```

```{r}
GDP_only %>% filter(GDP_pc != "NA") %>%
  group_by(continent) %>%
  summarize(avg_GDP_pc=mean(GDP_pc), 
            sd_GDP_pc=sd(GDP_pc), 
            min_GDP_pc=min(GDP_pc), 
            max_GDP_pc=max(GDP_pc),
            N=n())
```

```{r}
GDP_only %>% filter(GDP_pc != "NA") %>%
  filter(year==2010) %>%
  group_by(continent) %>%
  summarize(avg_GDP_pc=mean(GDP_pc), 
            sd_GDP_pc=sd(GDP_pc), 
            min_GDP_pc=min(GDP_pc), 
            max_GDP_pc=max(GDP_pc),
            N=n())
```

## How to view just a few observations

Challenge Question: Find out the top three GDP per capita in each continent.

- When you think you've got it figured out, remove the `eval=FALSE` option in the code chunk.

```{r, eval=FALSE}
___ <- GDP_only %>% filter(continent==___) %>% arrange(desc(GDP_pc))
head(___, ___)
```

## More practice: Gapminder

In teams of 2-3, answer the following questions.

1. Which countries have the highest infant mortality rates?
2. Which European counties have the highest life expectancy?
3. Which European countries have the lowest life expectancy?
4. Find the average infant mortality rate, worldwide.
5. Find the average infant mortality rate in each region.
6. Find the average infant mortality rate in each year.

