---
title: 'Lab 09: Making It Shiny'
subtitle: "MTH 365: Intro to Data Science"
author: "Your Name Here"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

This lab is designed to walk you through the steps of creating and deploying your own RShiny web applet. First, what the heck is Shiny?

__Shiny__ is an R package that makes it easy to build interactive web applications straight from R. 

As always, your first step should be to install the `shiny` package.

```{r}
# install.packages('shiny')
```

## Example: The simplest app ever...

Let's write your first Shiny app.

```{r}
library(shiny)
server <- function(input, output, session) {}
ui <- basicPage('This is my first Shiny app.')

shinyApp(ui=ui, server=server)
```

In the background, Shiny uses a web design framework called Twitter Bootstrap. This provides a basic layout and some minimal styling for your app.

Running the function `shinyApp()` essentially takes your code and compiles it to a web-friendly language, just like knitting an RMarkdown does. The difference between a Shiny app and a Markdown document is __interaction__. When you create a Shiny app, you're building an interactive data "dashboard". 

## Examples

There are a few built-in examples we can use to get an idea of the features Shiny provides.

```{r, eval=FALSE}
runExample('01_hello')
```

Here's the Shiny code required to build this example. There are two basic parts to any Shiny app:

- `ui`: This defines the layout and inputs for the `u`ser `i`nterface.
- `server`: This is where the R code that runs behind the Shiny app (on the `server`) needs to go.

```{r, eval=FALSE}
library(shiny)

# Define UI for app that draws a histogram ----
ui <- fluidPage(

  # App title ----
  titlePanel("Hello Shiny!"),

  # Sidebar layout with input and output definitions ----
  sidebarLayout(

    # Sidebar panel for inputs ----
    sidebarPanel(

      # Input: Slider for the number of bins ----
      sliderInput(inputId = "bins",
                  label = "Number of bins:",
                  min = 1,
                  max = 50,
                  value = 30)

    ),

    # Main panel for displaying outputs ----
    mainPanel(

      # Output: Histogram ----
      plotOutput(outputId = "distPlot")

    )
  )
)

# Define server logic required to draw a histogram ----
server <- function(input, output) {

  # Histogram of the Old Faithful Geyser Data ----
  # with requested number of bins
  # This expression that generates a histogram is wrapped in a call
  # to renderPlot to indicate that:
  #
  # 1. It is "reactive" and therefore should be automatically
  #    re-executed when inputs (input$bins) change
  # 2. Its output type is a plot
  output$distPlot <- renderPlot({

    x    <- faithful$waiting
    bins <- seq(min(x), max(x), length.out = input$bins + 1)

    hist(x, breaks = bins, col = "#75AADB", border = "white",
         xlab = "Waiting time to next eruption (in mins)",
         main = "Histogram of waiting times")

    })

}

# Create Shiny app ----
shinyApp(ui = ui, server = server)
```

Try some other examples! Run these one at a time, and see what they create.

- runExample("02_text")       
- runExample("03_reactivity")  
- runExample("04_mpg")         
- runExample("05_sliders")     
- runExample("06_tabsets")    
- runExample("07_widgets") 

```{r, eval=FALSE}
runExample("03_reactivity")
```

### Resources

- Zev Ross has an incredibly detailed reference guide for building Shiny apps: http://zevross.com/blog/2016/04/19/r-powered-web-applications-with-shiny-a-tutorial-and-cheat-sheet-with-40-example-apps/
- _Modern Data Science with R_: Ch. 11.3 "Interactive web apps with Shiny"

### Troubleshooting

You'll find instructions and tips in this lab document, however, you should NOT write your Shiny apps in a Markdown document. Why??

- Markdown documents compile as "static HTML" files - dynamic inputs aren't allowed. 
- However, you CAN run Shiny code in a Markdown document using the "play" button. It just won't compile correctly when you knit.

Instead, go to File -> New File -> Shiny Web App... to start building your app. 

## Tutorial: Exploring the `mtcars` dataset

The `mtcars` data set contains data from a 1974 issue of _Motor Trend_ magazine. Data on 10 aspects of automobile design and performance were collected for 32 automobiles.

Variables|Description
---|---
`mpg`| Miles per gallon
`cyl`|Number of cylinders
`disp`|Displacement (cubic inches)
`hp`|Gross horsepower
`drat`|Rear axle ratio
`wt`|Weight (in thousands of pounds)
`qsec`|1/4 mile time (in seconds)
`vs`|Engine (0 = v-shaped, 1 = straight)
`am`|Transmission (0 = automatic, 1 = manual)
`gear`|Number of forward gears
`carb`|Number of carburetors

Let's build a Shiny web app to visualize how miles per gallon is affected by various variables in this data set.

```{r}
data(mtcars)

hist(mtcars$mpg, col='cyan', main='Mileage Per Gallon')
```

> 1. Start by modifying the user interface.

- Update the `titlePanel` with a relevant title. 
- We'll start with a histogram. How many bins might be appropriate for the `mpg` variable? Modify the `sliderInput` to cover a range of 5-10 bin widths.

Run your app, and make sure the changes you've created work the way you want them to.

> 2. Next, change the data. 

Look for the `renderPlot()` function in the `server` logic. 

- Change `x` to take the `mpg` variable in the `mtcars` data set.
- Change the histogram color to a color of your choice.
- Add an axis label to your histogram (`xlab=_____`) and suppress the main title (`main=''`).

Note: The example Shiny template that you begin with in R uses "base R" instead of `tidyverse` to create plots. Since our goal for today is to get the hang of how Shiny works, and not create "publication-quality" visulations, that's okay. For Mini-Project 3, I'd like to see you use `ggplot` style graphics.

Run your app.

> 3. Let's add a second plot for horsepower.

- Copy the `output$distPlot` code segment, and paste it below your code (within the `server` call). Rename your output.
- Change the histogram variable to the `hp` variable in the `mtcars` data set. 
- Change the histogram color to a color of your choice, and modify the axis labels/title.
- Look for the `mainPanel()` function. Add your new plot to the `plotOutput` in this function.

```{r, eval=FALSE}
      # Show a plot of the generated distribution
      mainPanel(
         plotOutput("distPlot"),
         plotOutput(__________)
      )
```

Run your app.

> 4. Now let's improve the user interface. 

- Add a description of this data set in your sidebar.
- Move the sidebar to the right side of your app.

```{r, eval=FALSE}
   # Sidebar with a slider input for number of bins 
   sidebarLayout(position='right',
      sidebarPanel(
        "_____________________",
         sliderInput("bins",
                     "Number of bins:",
                     min = 2,
                     max = 10,
                     value = 5)
      ),
```

- Add a header and description before each plot in your main panel. The `h1()` and `h2()` functions create headers (as you increase the number, you decrease the size of the header), and the `p()` function creates plain text.

```{r, eval=FALSE}
      # Show a plot of the generated distribution
      mainPanel(h1(______),
                p(______),
                h2(______, align='center'),
                  plotOutput("distPlot"),
                  p(______),
                h2(______, align='center'),
                  plotOutput('distPlot2'),
                  p(______)
      )
```

Run your app.

A __widget__ is a web element that your users can interact with. You've got one already: a _slider_ that changes the number of bins. Let's add more.

> 5. Create a new widget that changes the color of your histogram.

- Choose any four colors to use as options in your select input widget.

__DO NOT__ run your app.

> 6. Create a new widget that allows users to select either cars with automatic transmissions only, cars with manual transmissions only, or the entire data set.

```{r, eval=FALSE}
   # Sidebar with a slider input for number of bins 
   sidebarLayout(position='right',
      sidebarPanel(
        "The `mtcars` data set contains data from a 1974 issue of _Motor Trend_ magazine. Data on 10 aspects of automobile design and performance wre collected for 32 automobiles.",
         sliderInput("bins",
                     "Number of bins:",
                     min = 2,
                     max = 10,
                     value = 5),
        
        selectInput("col", 
                    label = "Choose a plot color:",
                    choices = list(_____, 
                                   _____,
                                   _____, 
                                   _____),
                    selected = _____),
        
        selectInput("data", 
                    label = "Choose a subgroup:",
                    choices = list("Automatic transmissions",
                                   "Manual transmissions", 
                                   "All cars"),
                    selected = "All cars")
      ),
```

__DO NOT__ run your app.

> 7. Link your new widget inputs to the plot options in the `server` panel.

- There are now three `input` values: `bins`, `col`, and `data`.
- To link the plot color to your histogram, set the plotting color to `col = input$col`.
- Linking the data is harder. The sample code below will take the `data` input, and match it up with the subset of the data we want. Modify this for the horsepower.

```{r, eval=FALSE}
x <- switch(input$data, 
                "Automatic transmissions" = subset(mtcars, am==0)$mpg,
                "Manual transmissions" = subset(mtcars, am==1)$mpg,
                "All cars" = mtcars$mpg)
```

Run your app.

> 8. Finally, add a description on your app of how miles per gallon and horsepower change depending on engine type.'

Run your app.

------

When you're finished, upload `app.R` to BlueLine.

