---
title: "Lab 4: Scraping the Bottom of the (Beer) Barrel"
subtitle: "MTH 365: Intro to Data Science"
author: "Your Name Here"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

In 2017, the U.S. beer industry shipped (sold) 207.4 million barrels of beer – equivalent to more than 2.9 billion cases of 24-12 ounce servings. In addition, the industry shipped approximately 2 million barrels of cider, equivalent to more than 28.3 million cases. Additionally, the U.S. beer industry sells more than $111.1 billion in beer and malt-based beverages to U.S. consumers each year. (Source: [The U.S. Beer Industry]( https://www.nbwa.org/resources/industry-fast-facts))

Do you know how many breweries are around you? And how much beer they brew? In this lab we scrape and analyze data on US breweries.

We will start with getting data on breweries in Nebraska. Then, you will use a similar approach to get data on breweries in a different state of our choosing.

We'll use the Selector Gadget tool within a Chrome browser.

- http://selectorgadget.com/

In this example we'll work with the `tidyverse`, `rvest`, and `robotstxt` packages. You may need to install `rvest` and `robotstxt`.

```{r, message=FALSE, warning=FALSE}
library(tidyverse) 
library(rvest)
library(robotstxt)
```

# The data

We will scrape brewery information from https://www.ratebeer.com/breweries/. Ratebeer.com is an in-depth, consumer-driven source of beer information. We will use the state-level brewery lists on this site to first obtain information on all breweries in a given state. Then, we will dive deeper and obtain additional information on each of the breweries in that state, one-by-one, by automating our code to do so.

Before getting started, let's check that a bot has permissions to access pages on this domain.

```{r,warning=FALSE}
paths_allowed("https://www.ratebeer.com/")
```

## Nebraska breweries

Click the link to check out the Nebraksa page on ratebeer.com.

https://www.ratebeer.com/breweries/nebraska/27/213/

> 1. Based on the information on the Nebraska breweries page, how many total (active + closed) breweries are there in NE?

Make a new R scipt (not a Markdown document)! This is where we'll build our sample code.

- Let's start with the names of breweries. Use the selector gadget to determine the path to the names of breweries, and create a vector of type character which are the names of the breweries.

Next we get the cities. The paths for the cities on the active vs. closed page are different.

- Create two vectors, one called `active_cities` and the other `closed_cities` that contain the cities of the active and closed breweries, respectively. Then, combine these vectors with 

```{r, eval=FALSE}
cities <- c(active_cities, closed_cities)
```

- Scrape brewery type, number of beers brewed at each brewery, year when brewery first opened ("est.")s. Save these as vectors called `types`, `counts`, `ests`,  `urls`, and `status` respectively.

- Create a data frame (`tibble`) called `Nebraska_Beers` with column names shown in the table above.

- Save this data frame as `Nebraska_Beers.csv` using the `write_csv` function: `write.csv(Nebraska_Beers, path = "Nebraska_Beers.csv")`.

2. Load `Nebraska_Beers.csv` in your Rmd.

3. Which city in NE has the most breweries? How many breweries are in Omaha, NE? What are they? (Hint: There might be a few cities that need some more attention.)

4. Make a plot showing how the number of breweries in Nebraska has changed over time. Does the type of establishment play a role?


## Choose your own state

In a new R script, repeat the previous exercises (potentially with some modifications) to create a similar data frame (with the same columns) for a different state of your choosing. Save the result in a csc file with the name of the state you chose, e.g. `California_Beers.csv`.

Load the new data in the R Markdown.

6. Determine which city in that state has the highest number breweries.

7. Determine which city has the youngest breweries, on average.

8. Write a function to grab the zip code for each of the breweries from their own pages (at the URL you recorded). Test this function out on the first three breweries.

9. By using the `map` function, grab the zip code for all of the breweries from their own pages (at the URL you recorded) and add this as a new column to your data frame.

10. Determine which zip code in that state has the highest number breweries.

For your submission, include both the `script.R` file you used to scrape the data and your compiled RMarkdown lab.